---
title: 'Economic and population health impacts of NOAA storms'
output: html_document
---

## Synopsis

We'll address the following two questions:

1. Across the United States, which types of storm events are most harmful with respect
to population health?
2. Across the United States, which types of storm events have the greatest economic 
consequences?

To investigate, we obtained data from the NOAA storm database.  We tidy up the data
because the events were evidently entered in a free-form manner. From these data, 
we found that floods and hurricanes have the greatest economic consequence, and
that tornadoes and heat are the most harmful with respect to human health.


## Data Processing

### Obtaining the data
We obtained the [NOAA storm database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), downloading it to a local directory. For purposes of this 
assignment, we're asked to start from the the raw data file, not the url. In fact 
the assessment specifically calls for the analysis to start from the original 
.csv.bz2 file. For this reason, code for downloading the file is not shown.

### Reading in the data
R is able to read compressed archives directly, so the repdata-data-StormData.csv.bz2 
file is not decompressed first.  The file data is loaded into an R data frame. 
The stringsAsFactors argument is set to false because its arguably easier in R 
to perform cleanup work with strings. Its reasonable to assume at this early 
stage that at least some cleanup will occur.  Since the check.names arguments of 
the read.table() function defaults to true, the variable names will be 
syntically valid per R rules.


```{r, echo=TRUE,cache=TRUE}
ptm <- proc.time() # Start the clock
storms <- read.table("repdata-data-StormData.csv.bz2",sep=",",
                     stringsAsFactors=F,header=T,strip.white=T)
proc.time() - ptm # Stop the clock and display the elapsed time
```

After loading the data, we inspect it it using our R "toolbox".

```{r,echo=TRUE}
dim(storms)
head(storms)
str(storms)
evtype_cnt <- length(unique(storms$EVTYPE))
```

The columns we're interested in are the following, because they directly relate 
to our two questions involving event types with greatest impact.

Column Name   | Description
------------- | -------------
EVTYPE        | Storm Event Type
FATALITIES    | Number of Fatalities
INJURIES      | Number of Injuries
CROPDMG       | Crop damage units. Paired with the units abbreviation CROPDMGEXP 
CROPDMGEXP    | Crop damage unit abbreviation ( hundreds, thousands, millions or billions )
PROPDMG       | Property damage units.  Paired with the units abbreviation PROPDMGEXP 
PROPDMGEXP    | Property damage unit abbreviation ( hundreds, thousands, millions or billions )

### Transforming the data

The EVTYPE variable is critical for analysis since it contains the category of
severe weather event. There are `r evtype_cnt` distinct raw values in the EVTYPE 
variable. However, in the [NOAA storm database document](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) , there are 48 identified severe 
weather event categories.
Because the EVTYPE variable is both essential and dirty, we attempt to clean it 
up to match the standardized severe weather event categories.  The reason to clean 
up the EVTYPE column is to provide sufficient grouping to the free-form event 
types so we can get at the question of which event "types" are most damaging.

The EVTYPE cleanup involves lower-casing, digit removal, leading/trailing 
whitespace removal, and lastly, manual mapping. The reasons for adding all these
transformations are similar. After each transformation to EVTYPE, the number of 
unmatched storms drops. In terms of how much cleanup is good enough, we're 
aiming to categorize enough events so that at least the top storm category for 
both questions is known.

```{r,echo=T}
# create vector of standard event types.  We lowercase so we can compare more
# easily to our free-form storm dataset event types
standard_event_types <- tolower(c(
    "Astronomical Low Tide","Avalanche","Blizzard","Coastal Flood",
    "Cold/Wind Chill","Debris Flow","Dense Fog","Dense Smoke","Drought",
    "Dust Devil","Dust Storm","Excessive Heat","Extreme Cold/Wind Chill",
    "Flash Flood","Flood","Frost/Freeze","Funnel Cloud","Freezing Fog",
    "Hail","Heat","Heavy Rain","Heavy Snow","High Surf","High Wind",
    "Hurricane (Typhoon)","Ice Storm","Lake-Effect Snow","Lakeshore Flood",
    "Lightning","Marine Hail","Marine High Wind","Marine Strong Wind",
    "Marine Thunderstorm Wind","Rip Current","Seiche","Sleet","Storm Surge/Tide",
    "Strong Wind","Thunderstorm Wind","Tornado","Tropical Depression",
    "Tropical Storm","Tsunami","Volcanic Ash","Waterspout","Wildfire",
    "Winter Storm","Winter Weather"
))

storms$EVTYPE <- tolower(storms$EVTYPE) # Lower case
storms$EVTYPE <- gsub("\\d","",storms$EVTYPE) # Remove digits
# Removing leading and trailing whitespace http://stackoverflow.com/questions/2261079/how-to-trim-leading-and-trailing-whitespace-in-r
storms$EVTYPE <- gsub("^\\s+|\\s+$", "", storms$EVTYPE) # Remove trailing spaces
# Take the remaining unmatched events, eye-ball the ones with a large number of 
# events and hand code them to standard event names for those with obvious matches.
storms[storms$EVTYPE == "marine tstm wind","EVTYPE"] <-
    "marine thunderstorm wind"
storms[storms$EVTYPE == "tstm wind","EVTYPE"] <-
    "thunderstorm wind"
storms[storms$EVTYPE == "thunderstorm winds","EVTYPE"] <-
    "thunderstorm wind"
storms[storms$EVTYPE == "extreme heat","EVTYPE"] <-
    "heat"
storms[storms$EVTYPE == "heat wave","EVTYPE"] <-
    "heat"
storms[storms$EVTYPE == "extreme cold","EVTYPE"] <-
    "extreme cold/wind chill"

# Circling back after calculating the monetary damages below, the most economically
# damaging storms have EVTYPE values that don't match standard storms types. Because
# of their significance ecnomically, such storms are manually categorized here using
# a judgement call (e.g it seems reasonable to match "Storm Surge" to 
# "Storm Surge/Tide".
storms[storms$EVTYPE == "storm surge","EVTYPE"] <-
    tolower("Storm Surge/Tide")
storms[storms$EVTYPE == "hurricane/typhoon","EVTYPE"] <-
    tolower("Hurricane (Typhoon)")
storms[storms$EVTYPE == "hurricane","EVTYPE"] <-
    tolower("Hurricane (Typhoon)")
storms[storms$EVTYPE == "hurricane opal","EVTYPE"] <-
    tolower("Hurricane (Typhoon)")
storms[storms$EVTYPE == "heavy rain/severe weather","EVTYPE"] <-
    tolower("Heavy Rain")
storms[storms$EVTYPE == "river flood","EVTYPE"] <-
    tolower("flood")

# Create a STD_EVTYPE, which maps to NOAA standard storm type categories
# and uses "other" for un-matching storms.
storms$STD_EVTYPE <- ifelse(storms$EVTYPE %in% standard_event_types,
                            storms$EVTYPE,"other")

# Determine percentage of storms categorized
pct_categorized <- 100 - (prop.table(table(storms$STD_EVTYPE))*100)[["other"]]
```
We finally concluded with `r pct_categorized`% of the storms categorized, which
for purposes of this analysis, we'll assume to be a wide enough net to to capture
the most damaging of events, either in population health or monetary terms. We'll 
double check this assumption later on by checking the population health impact 
and monetary value of uncategorized events.

The CROPDMGEXP and PROPDMGEXP variables also contain dirty data. In page 12 of the 
documentation, values of "h" for hundreds, "k" for thousands, "m" for millions,
and "b" for billions are specified. The actual set of values is different
from this specification.  

```{r,echo=T,message=F}
# plyr is used later by treemapify(), or at least that was the idea. If both 
# plyr and dplyr are used they should be loaded in the following sequence.
library(reshape2)
library(plyr)
library(dplyr)
```

```{r,echo=T}
storms$STORM_COUNT <- 1
# storms %>% group_by(tolower(PROPDMGEXP)) %>% summarize(storm_count=sum(STORM_COUNT))
# storms %>% group_by(tolower(CROPDMGEXP)) %>% summarize(storm_count=sum(STORM_COUNT))
```

Since the number of events with nonstandard EXP values is small, for purposes of 
this analysis, these values will be assumed to be "single". "single" means that 
for those EXP observations not containg a standard unit abbreviation, the value 
of the corresponding DMG column will be multiplied by one. Here, we create 
PROPDMGAMT and CROPDMGAMT variables to contain calculated monetary damages. For 
example, PROPDMGAMT is based on the raw variables PROPDMG and PROPDMGEXP.  These 
monetary damages will be much easier to use than the raw multi-column values.

```{r,echo=TRUE}
storms$PROPDMGEXP <- tolower(storms$PROPDMGEXP)
storms$CROPDMGEXP <- tolower(storms$CROPDMGEXP)

# Create a list which maps between unit abbreviations and related powers of 10
unit_abbreviations <- list(h=2,k=3,m=6,b=9,single=0)

# Standardize *EXP columns to match unit list
storms$PROPDMGEXP <- 
    ifelse(storms$PROPDMGEXP %in% names(unit_abbreviations),storms$PROPDMGEXP,
           "single")
storms$CROPDMGEXP <- 
    ifelse(storms$CROPDMGEXP %in% names(unit_abbreviations),storms$CROPDMGEXP,
           "single")

# Lookup unit abbreviations and raise 10 this power, then multiply 
# by the related DMG column to arrive at monetary damages.
storms$PROPDMGAMT <-
    storms$PROPDMG * (10 ^ (unlist(unit_abbreviations[storms$PROPDMGEXP])))
storms$CROPDMGAMT <-
    storms$CROPDMG * (10 ^ (unlist(unit_abbreviations[storms$CROPDMGEXP])))

```

Now we have property and crop damages calculated. We then double check our 
assumption that we have categorized "enough" of the storms to find the most
damaging of events. 

```{r,echo=TRUE}
fatalities_by_type <- storms %>% group_by(STD_EVTYPE) %>% 
    summarize(fatalities = sum(FATALITIES))
fatalities_by_type <- mutate(fatalities_by_type,rnk = rank(-fatalities))

injuries_by_type <- storms %>% group_by(STD_EVTYPE) %>% 
    summarize(injuries = sum(INJURIES))
injuries_by_type <- mutate(injuries_by_type,rnk = rank(-injuries))

econ_dmg_by_type <- storms %>% group_by(STD_EVTYPE) %>% 
    summarize(damages = sum(PROPDMGAMT + CROPDMGAMT))
econ_dmg_by_type <- mutate(econ_dmg_by_type,rnk = rank(-damages))

# with(fatalities_by_type,head(fatalities_by_type[order(-fatalities),]))
uncategorized_fatalities_rnk <- with(fatalities_by_type,
                  fatalities_by_type[STD_EVTYPE=="other",c("rnk")])$rnk

# with(econ_dmg_by_type,head(econ_dmg_by_type[order(-damages),],8))
uncategorized_damages_rnk <- with(econ_dmg_by_type,
                  econ_dmg_by_type[STD_EVTYPE=="other",c("rnk")])$rnk

# with(injuries_by_type,head(injuries_by_type[order(-injuries),],10))
uncategorized_injuries_rnk <- with(injuries_by_type,
                  injuries_by_type[STD_EVTYPE=="other",c("rnk")])$rnk
```

In terms of fatalities, the "other" category, which represents uncategorized 
storms ranks `r uncategorized_fatalities_rnk` and for injuries, the "other" 
category ranks `r uncategorized_injuries_rnk`, so our assumption is a safe one 
to have made for the population health question.

In terms of economic damages, the "other" category, which represents uncategorized 
storms, initially ranked 1, so this was not a safe assumption to have made. 
We'll revisit the storm categorization. After categorizing highly damaging 
events with non-standard categorization, "other" category dropped to a 
rank of  `r uncategorized_damages_rnk`, so we now assume to have categorized 
enough events.

## Results

For purposes of determining relative "harm" to population health from a storm, 
we assume the FATALITIES variable as a proxy for harm. Some consideration was 
given to use of INJURIES variable as well, certainly an injury is harmful to 
population health.  However, the relative "harm" of a FATALITY vs an INJURY is 
difficult if not impossible to quantify given this data set.

Here we produce a scatterplot of economic damage and harm to population health.
Each point indicates the totals for a given standard EVTYPE.  Those points to
the right are storm event types with the greatest harm to population health. 
Those points to the top are event types with the greatest economic impact.

```{r,echo=T,fig.width=10,fig.height=6}
library(scales)
library(ggplot2)
storm_type_totals <- storms %>% group_by(STD_EVTYPE) %>% 
    summarize(fatalities = sum(FATALITIES),damages = sum(PROPDMGAMT + CROPDMGAMT))
p <- ggplot(data=storm_type_totals,aes(x=fatalities,y=damages))+
    geom_point()+
    geom_text(aes(label=STD_EVTYPE),size=4,vjust=1,hjust=0.25)+
    xlab("Fatalities")+
    scale_y_continuous(name="Economic Damages (USD)", labels = comma)+    
    ggtitle("Harm to Population and enconomic impact of Severe Storms")
print(p)
```

Across the United States, the types of event most harmful with respect
to population health are 1) tornadoes and 2) excessive heat. 

In terms of economic impact, we combined both property and crop damages. This
combination of the two separate economic impacts is our measure of relative 
economic impact.  From the above graph, we can see storm type with greatest 
economic impact is flooding, with an economic toll of around $160 billion. 
Hurricanes, and their associated storm surge, are the second most impactful 
storm type.

